Tracking Rodinia into the Neoproterozoic: new paleomagnetic constraints from the Jacobsville Formation

### 2024-01-23 Manuscript/Manuscript.tex
update manuscript for proof

\begin{keypoints}
\item Siltstones and fine-grained sandstones of the ca. 990 Ma Jacobsville Formation have primary magnetization held by detrital hematite.
\item Following closure of the Unimos Ocean and onset of the Grenvillian orogeny in the late Mesoproterozoic, Laurentia's motion significantly slowed.
\item The pole position and age suggests that the Grenville Loop paleomagnetic poles are younger than current interpretations.
\end{keypoints}

\begin{abstract}
The paleogeography of Laurentia throughout the Neoproterozoic is critical for reconstructing global paleogeography due to its central position in the supercontinent Rodinia. We develop a new paleomagnetic pole from red siltstones and fine-grained sandstones of the early Neoproterozoic Jacobsville Formation which is now constrained to be ca. 990 Ma in age. High-resolution thermal demagnetization experiments resolve detrital remanent magnetizations held by hematite. These directions were reoriented within siltstone rip-up clasts and pass an intraformational conglomerate test---giving confidence that the magnetization is detrital and primary. An inclination-corrected mean paleomagnetic pole position for the Jacobsville Formation indicates that Laurentia's motion slowed down significantly following the onset of the Grenvillian orogeny. Prior rapid plate motion associated with closure of the Unimos Ocean transitioned to slow drift of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. We interpret the distinct position of this well-dated pole from those in the Grenville orogen that have been assigned a similar age to indicate that the ages of the poles associated with the Grenville Loop likely need to be revised to be younger due to prolonged exhumation. 
\end{abstract}

@@

The extensively studied paleomagnetic records of ca. 1109 to 1084 Ma volcanics and intrusions associated with the North American Midcontinent Rift provide crucial constraints on the paleogeography of Laurentia in the late Mesoproterozoic Era \cite{Halls1982a, Fairchild2017a, Swanson-Hysell2019a}. The resulting sequence of poles known as the Keweenawan Track is a central record for reconstructing the assembly of the ancient supercontinent Rodinia \cite{Evans2021b}. An advantage of intracratonic magmatic events, such as those preserved in the Midcontinent Rift, is that they lead to emplacement of igneous rocks that can readily retain primary thermal remanent magnetization (TRM) that can be confidently associated with the emplacement age of the unit. However, following the development of the Midcontinent Rift, there was a $\sim$300 Myr quiescence in known intracratonic magmatic activity in Laurentia that lasted until the emplacement of the ca. 775 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a, Swanson-Hysell2021c}. The lack of Laurentian rocks with primary TRM in the early Neoproterozoic has led paleogeography reconstructions to be reliant on magnetizations acquired by rocks that were metamorphosed during the ca. 1090 to 980 Ma Grenvillian orogeny \cite{Rivers2012a, Swanson-Hysell2023a}. However, rocks within the orogen acquired their magnetizations during exhumation long after peak metamorphic conditions \cite{McWilliams1975a}. As a result, the timing of remanence acquisition needs to be constrained through the more difficult task of dating cooling associated with exhumation. This gap in well-dated paleomagnetic poles for Laurentia in the late Mesoproterozoic to mid-Neoproterozoic contributes to uncertainty in paleogeography at that time including the configuration of the supercontinent Rodinia.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing the extent of the Midcontinent Rift and the location of the Grenville Front relative to the study area. The inset green box shows the extent of panel B. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, the Jacobsville Formation, sedimentary rocks of the Bayfield Group, other older Proterozoic rocks, and major post-rift thrust faults. The location of the Jacobsville stratigraphic sections in this study are shown with green diamonds. Map modified from \citeA{Nicholson2004a}.}
\label{fig:Geologic_map}
\end{figure*}

@@

One rare preservation of an early Neoproterozoic sedimentary succession in Laurentia's interior is the Jacobsville Formation (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). A 992.51$\pm$0.64 Ma (2$\sigma$ analytical uncertainty) U-Pb detrital zircon date from the Jacobsville Formation constrains its maximum deposition age \cite{Hodgin2022a}. A 985.5$\pm$35.8 Ma (2$\sigma$ analytical uncertainty) U-Pb date from calcite veins within the Keweenaw fault provides a minimum age constraint on Jacobsville deposition, as the Jacobsville Formation is folded in the footwall against Midcontinent rift volcanics in the hanging wall. Together, these dates bracket the depositional age of the Jacobsville Formation and constrain it to be ca. 990 Ma \cite{Hodgin2022a}. These results constrain the formation to have been deposited within a syn-orogenic basin during the Rigolet phase of the Grenvillian orogen and to have been deformed during the same orogenic phase \cite{Hodgin2022a, Swanson-Hysell2023a}.

The Jacobsville Formation contains abundant hematite-bearing sandstone and siltstone which have the potential to record a key pole in Laurentia's apparent polar wander path (APWP) \cite{Dubois1962a, Roy1978a}. \citeA{Roy1978a} conducted a paleomagnetic study that developed a pole which has been used in paleogeographic syntheses where it has been assigned variable ages including an age of ca. 1020 Ma based on APWP extrapolation \cite<e.g.>{Li2008a}. However, the uncertainty in its age combined with a lack of field tests, and no application of inclination correction have resulted in the pole not being included in recent curated compilations \cite<e.g.>{Evans2021a}. Obtaining high-quality paleomagnetic data, including robust field tests, motivates the development of new data from the Jacobsville Formation.

In this study, we investigate five stratigraphic sections of the Jacobsville Formation on and near the Keweenaw Peninsula, Michigan, USA (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). High-resolution thermal demagnetization isolates primary detrital remanent magnetization (DRM) from secondary chemical remanent magnetization (CRM) within analyzed samples. This interpretation is supported by the DRM directions passing an intraformational conglomerate test on siltstone intraclasts that were ripped-up and remobilized within the fluvial depositional environment. A paleomagnetic fold test further constrains the timing of DRM acquisition to predate the folding of the Jacobsville Formation in the footwall of the Keweenaw Fault soon after deposition. With these data, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples that gives new constraints on the paleogeographic position of Laurentia in the earliest Neoproterozoic.

@@

The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). Following the end of active extension in the rift, thermal subsidence resulted in the accumulation of $>$4 km of ca. 1085-1050 Ma Oronto Group sedimentary rocks that are both intercalated with the youngest volcanics and deposited atop them (Fig. \ref{fig:Geologic_map}; \citeA{Daniels1982a, Cannon1989a}). Subsequently, the rift was inverted as contraction associated with the ca. 1090-980 Ma Grenvillian Orogeny on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}) propagated into Laurentia's interior \cite{Cannon1993a, Cannon1994a, Hodgin2022a, Swanson-Hysell2023a}.

During rift inversion, Midcontinent Rift volcanic and sedimentary rocks were uplifted along with Paleoproterozoic and Archean lithologies via thrust faults such as the Marenisco fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). Rb-Sr radiometric thermochronology data of \citeA{Cannon1993a} show Archean lithologies within the Montreal River monocline to have been exhumed to mid to shallow crustal temperatures ($\sim$270\textdegree C) by ca. 1050 Ma. The Jacobsville Formation overlies an erosional angular unconformity that developed on lithologies that were exhumed through this earlier episode of contractional deformation associated with Grenvillian orogenesis (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). Conglomeratic facies can occur in the basal Jacobsville Formation, with clasts derived from locally uplifted basement lithologies \cite{Irving1885a, Hamblin1958a, Kalliokoski1982a, Alemu2023a}.

Bedding planes in the Jacobsville Formation typically have shallow dips. The exception to these near-horizontal orientations is proximal to the Keweenaw fault (Fig. \ref{fig:Geologic_map}) where the Jacobsville Formation was occasionally deformed into monoclinal drag folds resulting in steeply tilted to overturned beds (Fig. \ref{fig:Geologic_map}; \citeA{Irving1885a, Cannon2001a}). Recent mapping of the Jacobsville Formation also found that it sometimes onlaps the Midcontinent Rift volcanics within or adjacent to en echelon fault branches of the Keweenaw fault system near the northern end of the Keweenaw Peninsula \cite{Tyrrell2019a, Mueller2021a}. Clumped isotope data for vein calcite with a U-Pb date of 985.5$\pm$35.8 Ma in the Keweenaw fault breccia within the hanging wall give temperatures of $\sim$50\textdegree C  \cite{Hodgin2022a}. The low temperature indicates that the calcite precipitated at shallow depth and that the majority of contractional inversion along the fault was complete by ca. 985 Ma. Combined with the 992.51$\pm$0.64 Ma maximum depositional age, these data are interpreted to indicate that the Jacobsville is a syn-orogenic unit deposited during the ca. 1010-980 Ma Rigolet phase of the Grenvillian orogeny \cite{Hodgin2022a, Swanson-Hysell2023a}.

In addition, field observations reveal erosional contacts and development of paleosols between the Jacobsville Formation and underlying volcanic rocks \cite{Hamblin1958a, Kalliokoski1975a}, and seismic reflection data beneath Lake Superior have been interpreted to indicate that the Jacobsville Formation lies in angular unconformity atop sedimentary rocks of the Oronto Group \cite{Cannon1989a}. These data indicate that there was an interval of erosion following ca. 1050 Ma prior to the onset of Jacobsville Formation deposition.

@@

The Jacobsville Formation is a fluvial sequence of feldspathic and quartzose conglomerates, sandstones, siltstones, and shales devoid of lava flows or cross-cutting igneous dikes although internal clastic dikes are present \cite{Hamblin1958a}. Sandstones of the Jacobsville Formation are more quartz-rich with fewer lithic fragments than sandstones of the Oronto Group such that they can be classified as subarkose to quartz arenites (Fig. \ref{fig:Field_photo}; \citeA{Hamblin1958a, Ojakangas2002a}). While much of the Jacobsville Formation is dominated by sandstone channel deposits, there are clay-rich siltstones that formed in overbank settings that are a major focus of this study.

\begin{figure*}[h!]
\centering
\includegraphics[width=0.8\textwidth]{Jacobsville_Sections_v6.pdf}
\caption{Lithostratigraphy and magnetostratigraphy for studied sections of the Jacobsville Formation in northern Michigan, USA. Paleomagnetic sampling focused on the dark red siltstone to fine-grained sandstone lithofacies. Paleomagnetic specimen declinations are plotted with blue circles corresponding to normal polarity and orange circles to reverse polarity. The maximum depositional age of the Agate Falls section is constrained by a CA-ID-TIMS detrital zircon U-Pb date of 1003.2 $\pm$ 2.2 Ma (sample AF1-29.3 of \citeA{Hodgin2022a}). In the Snake Creek tributary and Natural Wall sections, the position of the detrital zircon samples BSC1-92.5 and NW1–61.5 from \citeA{Hodgin2022a} are shown. No U-Pb zircon dates from these samples were younger than the age of Midcontinent Rift volcanics. A correlation of a facies association consisting of volcanic-clast conglomerate, coarse arkosic sandstone, and dark red clay-rich siltstone is inferred between the sections at Hungarian Falls (along Dover Creek and a tributary falls), Hammel Creek, and Natural Wall \cite{Brojanigo1984a}. The stratigraphic correlation between Jacobsville sections at Agate Falls and Snake Creek tributary with these correlated sections from the central Keweenaw Peninsula is relatively uncertain.}
\label{fig:strat_column}
\end{figure*}

In southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Marenisco fault and Pelton Creek fault is interpreted to represent syn-depositional development of local relief (Fig. \ref{fig:Geologic_map}; \citeA{Kalliokoski1982a, Hedgman2014a}). Close to the Keweenaw fault in the central Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}), the Jacobsville Formation consists dominantly of quartz-rich, trough cross-bedded, medium-grained sandstone, with locally abundant clast-supported pebble to cobble conglomerate, and interbeds of red, hematite-bearing, micaceous siltstone to fine-grained sandstone (Hungarian Falls, Hammel Creek, Natural Wall; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). In this region, the conglomerate contains abundant rounded to angular volcanic clasts as large as boulders that are likely derived from uplifted Midcontinent Rift volcanics \cite{Irving1885a, Brojanigo1984a}. Thicker intervals of dark red to brick red siltstone and very fine-grained sandstone are interbeded with coarser-grained sandstone and conglomerate at Agate Falls and Hungarian Falls (Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). Rip-up clasts of the siltstones are found within some channelized interbeds of sandstone and conglomerate (Fig. \ref{fig:intraclast_pmag}).

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of the Jacobsville Formation. (A) The red/white medium-grained sandstone with decimeter-scale trough cross-stratification in this image is a wide-spread lithology in the formation, but was not targeted for paleomagnetic sampling in this study. (B, C, D) At the Snake Creek tributary and the Natural Wall ravine in the northeastern Keweenaw Peninsula, intervals of red fine-grained sandstone to siltstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw fault and also the nearly horizontal upper strata. At the Hungarian Falls section (E), relatively thick red siltstone horizons interbedded or interfingering with sandstones and conglomerates are found along a waterfall associated with a stream that is a tributary to Dover Creek. Abundant reoriented siltstone rip-up clasts can be present within the conglomeratic layers above siltstone beds (Fig. \ref{fig:intraclast_pmag}). (F) At Agate Falls, nearly flat-lying red fissile siltstone beds are exposed. Subrounded detrital mica grains deposited parallel to the bedding plane are often present within the siltstone facies. Another set of siltstone rip-up clasts were sampled from a conglomeratic layer at Agate Falls for a paleomagnetic conglomerate test (Fig. \ref{fig:intraclast_pmag}).}
\label{fig:Field_photo}
\end{figure*}

The abundance of rift volcanic-clasts in the conglomeratic facies at Hungarian Falls, Hammel Creek, and Natural Wall indicates the presence of uplifted volcanics in proximity to the Keweenaw fault in the central part of the peninsula during Jacobsville deposition (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Coarse-grained arkosic to lithic sandstone is a facies associated with the volcanic-clast conglomerates and likely derived from mechanical weathering of the exhumed volcanics. Chemical weathering of exhumed volcanics from the hanging wall is a likely source of clay within the abundant fine-grained lithologies that are associated with the volcanic-clast conglomeratic facies \cite{Hamblin1958a}. Detrital mica grains up to $\sim$5 mm in size commonly found within siltstone to fine-grained sandstone are likely sourced from Paleoproterozoic and Archean lithologies in the region (Fig. \ref{fig:Geologic_map}).

The thickness of the Jacobsville strata can vary significantly in northern Michigan \cite{Hamblin1958a, Kalliokoski1982a}, and complete stratigraphic sections are not exposed at the studied localities (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Clay-rich red siltstone is more abundant in conjunction with conglomerates (e.g. 20-120 m in the Hungarian Falls section; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}) than with the cross-stratified sandstone which dominates much of the formation. Relatively thin lithic to arkosic sandstone beds, which are otherwise uncommon in the Jacobsville Formation, are commonly associated with the volcanic conglomerates and red siltstones near the Keweenaw Fault (Fig. \ref{fig:strat_column}). Together, the volcanic conglomerates, red siltstones, and arkosic sandstones make up a unique facies association. As previously interpreted by \citeA{Brojanigo1984a}, this facies association was likely deposited proximal to actively uplifting volcanics near the Hungarian Falls, Hammel Creek, and Natural Wall sections (Fig. \ref{fig:strat_column}). The relative stratigraphic positions of the Agate Falls and Snake Creek tributary sections are less certain (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}).

The Jacobsville Formation has been mapped to have a continuous extent from the Wisconsin and Michigan border northeast to the Keweenaw Peninsula and east to the Marquette region (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1995a, Cannon1996a, Cannon2001a}). In the eastern Lake Superior region near Sault Ste Marie, Ontario, there are similar exposures of sedimentary rocks that are also mapped as the Jacobsville Formation \cite{Hamblin1958a}. As observed on the Keweenaw Peninsula, these sandstones are within the footwall of faults where they are overthrusted by Midcontinent Rift volcanics near Mamainse Point \cite{Manson1994a}. To the west, the Jacobsville Formation has been interpreted to be correlative with sedimentary rocks of the Bayfield Group in northern Wisconsin and the Fond du Lac Formation and Hinckley Formation in Minnesota \cite{Hamblin1958a, Kalliokoski1982a, Ojakangas2001b}. A challenge associated with these correlations is that the ca. 1075 to 1050 Ma Freda Formation of the Oronto Group and the ca. 990 Ma Jacobsville Formation have quite similar fluvial lithofacies despite being deposited in distinct basinal settings at different times.

>>>>

\begin{keypoints}
\item Siltstones and fine-grained sandstones of the ca. 990 Ma Jacobsville Formation have primary magnetization held by detrital hematite.
\item Following closure of the Unimos Ocean and onset of the Grenvillian orogeny in the late Mesoproterozoic, Laurentia's motion went from fast to slow.
\item The pole position and age suggests that the paleomagnetic poles of the Grenville Loop are younger than in previous interpretations.
\end{keypoints}

\begin{abstract}
The paleogeography of Laurentia throughout the Neoproterozoic is critical for reconstructing global paleogeography due to its central position in the supercontinent Rodinia. We develop a new paleomagnetic pole from red siltstones and fine-grained sandstones of the early Neoproterozoic Jacobsville Formation which is now constrained to be ca. 990 Ma in age. High-resolution thermal demagnetization experiments resolve detrital remanent magnetizations held by hematite. These directions were reoriented within siltstone intraclasts and pass intraformational conglomerate tests---giving confidence that the magnetization is detrital and primary. An inclination-corrected mean paleomagnetic pole position for the Jacobsville Formation indicates that Laurentia's motion slowed down significantly following the onset of the Grenvillian orogeny. Prior rapid plate motion associated with closure of the Unimos Ocean between 1110 and 1090 Ma transitioned to slow drift of Laurentia across the equator in the late Mesoproterozoic to early Neoproterozoic. We interpret the distinct position of this well-dated pole from those in the Grenville orogen that have been assigned a similar age to indicate that the ages of the poles associated with the Grenville Loop likely need to be revised to be younger due to prolonged exhumation. 
\end{abstract}

@@

The extensively studied paleomagnetic records of ca. 1109 to 1084 Ma volcanics and intrusions associated with the North American Midcontinent Rift provide crucial constraints on the paleogeography of Laurentia in the late Mesoproterozoic Era \cite{Halls1982a, Fairchild2017a, Swanson-Hysell2019a}. The resulting sequence of poles known as the Keweenawan Track is a central record for reconstructing the assembly of the ancient supercontinent Rodinia \cite{Evans2021b}. An advantage of intracratonic magmatic events, such as those preserved in the Midcontinent Rift, is that they lead to emplacement of igneous rocks that can readily retain primary thermal remanent magnetization (TRM) that can be confidently associated with the emplacement age of the unit. However, following the development of the Midcontinent Rift, there was a $\sim$300 Myr quiescence in known intracratonic magmatic activity in Laurentia that lasted until the emplacement of the ca. 775 Ma Gunbarrel large igneous province \cite{Harlan2003a, Mackinder2019a, Swanson-Hysell2021c}. The lack of Laurentian rocks with primary TRM in the early Neoproterozoic has led paleogeographic reconstructions to be reliant on magnetizations acquired by rocks that were metamorphosed during the ca. 1090 to 980 Ma Grenvillian orogeny \cite{Rivers2008a, Rivers2012a, Swanson-Hysell2023a}. However, rocks within the orogen acquired their magnetizations during exhumation after peak metamorphic conditions \cite{McWilliams1975a}. As a result, the timing of remanence acquisition needs to be constrained through the more difficult task of dating cooling associated with exhumation. This gap in well-dated paleomagnetic poles for Laurentia in the late Mesoproterozoic to mid-Neoproterozoic contributes to uncertainty in paleogeography at that time including the configuration of the supercontinent Rodinia.

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Geologic_map.pdf}
\caption{(A) Regional map showing the extent of the Midcontinent Rift and the location of the Grenville Front relative to the study area. The inset green box shows the extent of panel B. (B) Geologic map of northern Michigan and Wisconsin showing Midcontinent Rift igneous and sedimentary rocks, sedimentary rocks of the Jacobsville Formation and Bayfield Group, other older Paleoproterozoic and Archean rocks, and major post-rift thrust faults. The location of the Jacobsville stratigraphic sections in this study are shown with green diamonds. Map modified from \citeA{Nicholson2004a}.}
\label{fig:Geologic_map}
\end{figure*}

@@

One rare preservation of an early Neoproterozoic sedimentary succession in Laurentia's interior is the Jacobsville Formation (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Hodgin2022a, DeGraff2022a}). U-Pb detrital zircon dates developed through chemical abrasion–isotope dilution–thermal ionization mass spectrometry (CA-ID-TIMS) of 1003.21 $\pm$ 2.23 Ma and 992.51$\pm$0.64 Ma (2$\sigma$ analytical uncertainty) constrain the maximum deposition age of the Jacobsville Formation \cite{Hodgin2022a}. A 985.5$\pm$35.8 Ma (2$\sigma$ analytical uncertainty) U-Pb date from calcite veins within the Keweenaw fault provides a minimum age constraint on Jacobsville deposition, as the Jacobsville Formation is folded in the footwall against Midcontinent rift volcanics in the hanging wall. Together, these dates bracket the depositional age of the Jacobsville Formation and constrain it to be ca. 990 Ma \cite{Hodgin2022a}. These results support the interpretation that the formation was deposited within a syn-orogenic basin during the Rigolet phase of the Grenvillian orogen and was then deformed during the same orogenic phase.

The Jacobsville Formation contains abundant hematite-bearing sandstone and siltstone which have the potential to record a key pole in Laurentia's apparent polar wander path (APWP) \cite{Dubois1962a, Roy1978a}. \citeA{Roy1978a} conducted a paleomagnetic study that developed a pole which has been used in paleogeographic syntheses where it has been assigned variable ages including an age of ca. 1020 Ma based on APWP extrapolation \cite<e.g.>{Li2008a}. However, the uncertainty in the age of the magnetization combined with a lack of field tests, and no application of inclination correction have resulted in the pole not being included in recent curated compilations \cite<e.g.>{Evans2021a}. Obtaining high-quality paleomagnetic data, including robust field tests, motivates the development of new data from the Jacobsville Formation.

In this study, we investigate five stratigraphic sections of the Jacobsville Formation on and near the Keweenaw Peninsula, Michigan, USA (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). These sections were chosen given the abundance of fine-grained siliciclastic lithologies where magnetization can be constrained through intraclast conglomerate field tests. Additionally, they are within the continuous bedrock belt from which chronostratigraphic constraints have been developed. In contrast to fluvial sandstones in other exposures such as in the eastern Lake Superior region near Sault Ste. Marie, the maximum and minimum depositional ages of these sections can be more confidently constrained.

Within sampled lithologies, high-resolution thermal demagnetization isolates primary detrital remanent magnetization (DRM) from secondary chemical remanent magnetization (CRM). This interpretation is supported by the DRM directions passing intraformational conglomerate tests on siltstone intraclasts that were ripped-up and remobilized within the fluvial depositional environment. A paleomagnetic fold test further constrains the timing of DRM acquisition to predate the folding of the Jacobsville Formation in the footwall of the Keweenaw Fault soon after deposition. With these data, we develop a new inclination-corrected paleomagnetic pole for the Jacobsville Formation based on a large number of samples that gives new constraints on the paleogeographic position of Laurentia in the earliest Neoproterozoic.

@@

The ca. 1109 Ma to 1084 Ma North American Midcontinent Rift is a major intracontinental rift system that extends over 2000 km through the Laurentia craton (Fig. \ref{fig:Geologic_map}). Following the end of active extension in the rift, thermal subsidence resulted in the accumulation of $>$4 km of ca. 1085-1050 Ma Oronto Group strata that are both intercalated with the youngest volcanics and deposited atop them (Fig. \ref{fig:Geologic_map}; \citeA{Daniels1982a, Cannon1989a}). Subsequently, the rift was inverted as contraction associated with the ca. 1090-980 Ma Grenvillian Orogeny on the eastern margin of Laurentia (Fig. \ref{fig:Geologic_map}) propagated into Laurentia's interior \cite{Cannon1993a, Cannon1994a, Hodgin2022a, Swanson-Hysell2023a}.

During rift inversion, Midcontinent Rift volcanic and sedimentary rocks were uplifted along with Paleoproterozoic and Archean lithologies via thrust faults such as the Marenisco fault, forming the crustal-scale Montreal River monocline (Fig. \ref{fig:Geologic_map}; \citeA{Cannon1993a}). Rb-Sr biotite thermochronology data of \citeA{Cannon1993a} show Archean lithologies within the Montreal River monocline to have been exhumed to mid to shallow crustal temperatures ($\sim$270\textdegree C) by ca. 1050 Ma. The Jacobsville Formation overlies an angular unconformity that developed on lithologies that were exhumed through this earlier episode of contractional deformation associated with Grenvillian orogenesis (Fig. \ref{fig:Geologic_map}; \citeA{Hamblin1958a, Cannon1993a, Kalliokoski1982a}). Conglomeratic facies occur in the basal Jacobsville Formation, with clasts derived from locally uplifted basement lithologies \cite{Irving1885a, Hamblin1958a, Kalliokoski1982a}.

Bedding planes in the Jacobsville Formation typically have shallow dips. The exception to these near-horizontal orientations is proximal to reverse faults such as the Keweenaw fault (Fig. \ref{fig:Geologic_map}) where the Jacobsville Formation was occasionally deformed into monoclinal drag folds resulting in steeply tilted to overturned beds (Fig. \ref{fig:Geologic_map}; \citeA{Irving1885a, Cannon2001a}). Recent mapping of the Jacobsville Formation also found that it sometimes onlaps the Midcontinent Rift volcanics within or adjacent to en echelon fault branches of the Keweenaw fault system near the northern end of the Keweenaw Peninsula \cite{Tyrrell2019a, Mueller2021a}. These observations are consistent with the interpretation that there was deposition of Jacobsville Formation sediments during faulting.

@@

The Jacobsville Formation is a fluvial succession of feldspathic and quartzose conglomerates, sandstones, siltstones, and shales devoid of lava flows or cross-cutting igneous dikes although internal clastic dikes are present \cite{Hamblin1958a}. Sandstones of the Jacobsville Formation are more quartz-rich with generally fewer lithic fragments than sandstones of the Oronto Group such that they can typically be classified as subarkose to quartz arenites (Fig. \ref{fig:Field_photo}; \citeA{Hamblin1958a, Ojakangas2002a}). While much of the Jacobsville Formation is dominated by sandstone channel deposits, there are clay-rich siltstones that formed in overbank settings that are a major focus of this study.

\begin{figure*}[h!]
\centering
\includegraphics[width=0.75\textwidth]{Jacobsville_Sections_v6.pdf}
\caption{Lithostratigraphy and magnetostratigraphy for studied sections of the Jacobsville Formation in northern Michigan, USA. GPS locations for the base and top of the sections are noted. Paleomagnetic sampling focused on the dark red siltstone to fine-grained sandstone lithofacies. Paleomagnetic specimen declinations are plotted with blue circles corresponding to normal polarity and orange circles to reverse polarity. The maximum depositional age of the Agate Falls section is constrained by a CA-ID-TIMS detrital zircon U-Pb date of 1003.2 $\pm$ 2.2 Ma (sample AF1-29.3 of \citeA{Hodgin2022a}). In the Snake Creek tributary and Natural Wall sections, the position of the detrital zircon samples BSC1-92.5 and NW1–61.5 from \citeA{Hodgin2022a} are shown. No U-Pb zircon dates from these samples were younger than the age of Midcontinent Rift volcanics. A correlation of a facies association consisting of volcanic-clast conglomerate, coarse arkosic sandstone, and dark red clay-rich siltstone is inferred between the sections at Dover Creek (along Dover Creek and a tributary falls), Hammel Creek, and Natural Wall \cite{Brojanigo1984a}. The stratigraphic correlation between Jacobsville sections at Agate Falls and Snake Creek tributary with these correlated sections from the central Keweenaw Peninsula is relatively uncertain.}
\label{fig:strat_column}
\end{figure*}

In southern exposures of the Jacobsville Formation, conglomerate facies typically have provenance sourced from Paleoproterozoic lithologies such as vein quartz and iron formation clasts \cite{Hamblin1958a, Kalliokoski1982a}. The occurrence of these conglomerates in proximity to the Marenisco fault and Pelton Creek fault and the downsection steepening of stratal dips interpreted as growth strata may represent syn-depositional development of local relief (Fig. \ref{fig:Geologic_map}; \citeA{Kalliokoski1982a, Hedgman1992a}). Close to the Keweenaw fault in the central Keweenaw Peninsula (Fig. \ref{fig:Geologic_map}), the Jacobsville Formation consists dominantly of quartz-rich, trough cross-bedded, medium-grained sandstone, with locally abundant clast-supported pebble to cobble conglomerate, and interbeds of red, hematite-bearing, micaceous siltstone to fine-grained sandstone (Dover Creek, Hammel Creek, Natural Wall; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). In this region, the conglomerate contains abundant rounded to angular volcanic clasts as large as boulders that are likely derived from uplifted Midcontinent Rift volcanics \cite{Irving1885a, Brojanigo1984a}. Thicker intervals of dark red to brick red siltstone and very fine-grained sandstone are interbedded with coarser-grained sandstone and conglomerate at Agate Falls and Dover Creek (Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}). Intraclasts of the siltstones are found within some channelized interbeds of sandstone and conglomerate (Fig. \ref{fig:intraclast_pmag}).

\begin{figure*}[h!]
\centering
\includegraphics[width=\textwidth]{Field_photo.png}
\caption{\scriptsize Field photos of the Jacobsville Formation. (A) The red/white medium-grained sandstone with decimeter-scale trough cross-stratification in this image (taken at the shore near the unincorporated community of Jacobsville 46.9819\textdegree N, 88.4068\textdegree W) is a wide-spread lithofacies in the formation, but was not targeted for paleomagnetic sampling in this study. (B, C, D) At the Snake Creek tributary and the Natural Wall ravine in the northeastern Keweenaw Peninsula, intervals of red fine-grained sandstone to siltstone beds can be found through the steeply dipping to overturned basal strata near the Keweenaw fault and also the nearly horizontal upper strata. At the Dover Creek section (E), thick red siltstone horizons interbedded or interfingering with sandstones and conglomerates are found along a Dover Creek tributary waterfall (47.1708\textdegree N, 88.4466\textdegree W; this waterfall is distinct, but close to Hungarian Falls). Abundant reoriented siltstone intraclasts can be present within the conglomeratic layers above siltstone beds (Fig. \ref{fig:intraclast_pmag}). (F) At Agate Falls, nearly flat-lying red fissile siltstone beds are exposed. Detrital mica grains deposited parallel to the bedding plane are often present within the siltstone facies. Another set of siltstone intraclasts were sampled from a conglomeratic layer at Agate Falls for a paleomagnetic conglomerate test (Fig. \ref{fig:intraclast_pmag}).}
\label{fig:Field_photo}
\end{figure*}

The abundance of clasts of rift volcanics in the conglomeratic facies at Dover Creek, Hammel Creek, and Natural Wall indicates the presence of uplifted volcanics in proximity to the Keweenaw fault in the central part of the peninsula during Jacobsville deposition (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Coarse-grained arkosic to lithic sandstone is a facies associated with the volcanic-clast conglomerates and likely derived from mechanical weathering of the exhumed volcanics. Chemical weathering of exhumed volcanics from the hanging wall is a likely source of clay within the abundant fine-grained lithologies that are associated with the volcanic-clast conglomeratic facies (Fig. \ref{fig:strat_column}. Detrital zircon provenance data from Jacobsville samples on the Keweenaw Peninsula developed by \citeA{Malone2020a} reveal zircon with dates overlapping with Midcontinent Rift volcanism consistent with this provenance interpretation.  Detrital mica grains up to $\sim$5 mm in size that are commonly found within siltstone to fine-grained sandstone within the Jacobsville are likely sourced from Paleoproterozoic and Archean lithologies in the region (Fig. \ref{fig:Geologic_map}) and preferentially settled in lower energy overbank settings.

The thickness of the Jacobsville strata can vary in northern Michigan \cite{Hamblin1958a, Kalliokoski1982a}, and complete stratigraphic sections are not exposed at the studied localities (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}). Clay-rich red siltstone is more abundant in conjunction with conglomerates (e.g. 20-120 m in the Dover Creek section; Figs. \ref{fig:strat_column} and \ref{fig:Field_photo}) than with the cross-stratified sandstone which dominates much of the formation. Relatively thin lithic to arkosic sandstone beds, which are otherwise uncommon in the Jacobsville Formation, are commonly associated with the volcanic conglomerates and red siltstones near the Keweenaw Fault (Fig. \ref{fig:strat_column}). Together, the volcanic conglomerates, red siltstones, and arkosic sandstones make up a unique facies association. As previously interpreted by \citeA{Brojanigo1984a}, this facies association was likely deposited proximal to actively uplifting volcanic rocks near the Dover Creek, Hammel Creek, and Natural Wall sections (Fig. \ref{fig:strat_column}). The relative stratigraphic positions of the Agate Falls and Snake Creek tributary sections are less certain (Figs. \ref{fig:Geologic_map} and \ref{fig:strat_column}).
